Two Decades of Smalltalk VM Development

### 2018-08-18 paper.tex
More typo correction.  Use \emph for all code. inor tweaks to language. Add a footnote detauling the amazingly meta experience of adding socket primitive support to the VM.

\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g. Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler.

@@

%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler within it. %We mention some of the limitations and how we worked around them. 
%We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. 
Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

%Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
In this section, we first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure. Then, we describe the memory layout of both the production and simulation runtimes. Next, we detail how VM execution is simulated and the interactions with the simulated memory.
In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator. The last subsection discusses simulation features related to the development of the JIT.

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code (interpreter, JIT and GC code) to C code, generating a few C files. This first step usually takes several seconds. Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable. This second step can take up to a few minutes the first time, but usually only takes a few seconds, since object files for unchanged C files do not need to be recreated.

The VM can be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is the most widely used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.

@@

\paragraph{Snapshots.}
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods from objects. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
. A new snapshot can be made during or at the end of the development session.

Snapshots can be used to avoid long start-up times when fixing specific bugs. The VM can be run to a point where the bug is about to be encountered and the system can be snapshotted at this point. Then multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.

@@

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the prducton VM and its simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

At higher addresses we find the beginning of the memory managed by the VM. At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

Usually at a very high addresse we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices inside the byte array instead of pointers. All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects. They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a doubly linked cycle of stack pages which are maintained by the VM. In the interpreter-only simulator, the stack zone is a Smalltalk Array object. In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.

@@

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add modularity through polymorphism. For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the modularity is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be reused for debugging purposes. Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

@@

In addition to the interpreter simulator, simulating the JIT requires simulating execution of the native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. When the JIT is enabled, the start of the byte array representing the memory is used to hold the machine code it generates at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through %Smalltalk primitives.  These are effectively a combination of an 
FFI calls and surrounding glue code.
Each processor simulator is invoked via calls that supply the byte array as a parameter, and effectively the byte array is the processor simulator's memory, while the processor simulator contains its register state which is accessed via Smalltalk's support for external memory access. \footnote{http://www.squeaksource.com/Alien}
% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation. %, and the debugging facilities for machine code that are available.

@@

The interpreter has \emph{framePointer}, \emph{stackPointer} and \emph{instructionPointer} variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the \emph{stackPointer} and \emph{framePointer} variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text, and all have fixed addresses. Referencing these variables and invoking these runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference each variable or routine, and uses this as the key in the relevant map, creating and storing a suitable closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is always invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, \emph{handleSimulationTrap:}, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.

@@

There is a small polymorphic scheme for breakpoints.  Booleans, integers, and arrays understand \emph{isBreakpointFor: pc}.  So if the \emph{breakPC} variable is \emph{false} there is no breakpoint.  If \emph{breakPC} is \emph{true} then any pc is a potential breakpoint.  If it is an integer then that pc is a potential breakpoint.  If the pc is an \emph{Array} then the pc is a breakpoint if any value in the array answers true to \emph{isBreakpointFor:}. The scheme could trivially be extended to include intervals. If \emph{breakPC isBreakpointFor: pc} is \emph{true} then the \emph{breakBlock} is sent \emph{shouldStopIfAtPC: pc}.  booleans and closures understand \emph{shouldStopIfAtPC:}.  \emph{true} and \emph{false} make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.  For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.  It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).

@@

Simulating the whole VM requires going through the entire Smalltalk system start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. In the simulator start-up takes around 15 seconds on a recent MacBook Pro. While developing the JIT, this start-up time may still be too long and degrade the live programming experience to an edit-compile-run cycle, which we prefer to avoid. In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part. This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code, decorates the disassembly and then dumps the output in a text window.

@@

\paragraph{Decorated Disassembly.}
Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in that disassembly and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the disassembly much more readable than that displayed in a typical machine oriented debugger.  For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, a check which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen. %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

@@

In this section we describe our personal experiences using the simulation infrastructure to develop the VM. The different subsections consist in different tasks we had to implement and in each case we emphasise how we used the simulator and how it helped us perform the each given task.

\subsection{Initial JIT development and new back ends}
When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable pattern.  As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the MessageNotUnderstood exception. The programmer would implement the body of the method and execution would continue.  Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good, somewhat like lemming debugging above.  As a new method was encountered and JITted yet more templates would be encountered.  Similarly as the abstract machine instruction to concrete machine instruction mappings were added these could be implemented in the debugger, an experience repeated when new back ends were added. Much of the JIT was written in this interactive live style.  This style encourages tool creation because one can create, extend and polish tools in the middle of debugging sessions, enhancing them to make sense of ones current predicament.  Immediate feedback makes tool investment cheap and cheerful.

@@

Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.  But because the simulation is simply an object graph it is easy to copy the entire simulation. Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).  Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.  We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging%\footnote{http://www.romainbrasier.fr/404/}
.  One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.  This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.

Recently, to evaluate some new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results. Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was working out of the box. %To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

@@

%In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version. This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap. Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each object at the exact same addresses. Indeed, in most cases when a bug happens, the memory was already corrupted and it was very difficult to track back where the bug came from without rewinding the memory state. We reproduced the exact same GC many times in different cases to fully understand specific bugs.

@@

One of the first analyses we ran was on the machine code zone. We stopped the simulation when the machine code zone reached 1Mb. We then iterated over it and investigated what was in it. As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).

@@

Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.

>>>>

\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g. Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.  Development is done in Smalltalk, an activity we call ``Simulation''. The production VM is derived by translating the core VM code to C. As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to an executable through a C compiler.

@@

%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler (JIT) within it. %We mention some of the limitations and how we worked around them. 
%We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. 
Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.

@@

%Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
In this section we first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure. Then we describe the memory layout of both the production and simulation runtimes. Next we detail how VM execution is simulated and the interactions with the simulated memory.
In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator. The last subsection discusses simulation features related to the development of the JIT.

@@

The VM executable is generated in a two step process. Firstly, the Slang-to-C compiler translates the Slang code (interpreter, JIT and GC code) to C code, generating a few C files. This first step usually takes several seconds. Secondly, the C compiler (depending on the platform LLVM, GCC or MSVC) translates all the C files (generated C files and platform-specific C files) into an executable. This second step can take up to a few minutes the first time, but usually only takes a few seconds, since object files for unchanged C files do not need to be recreated.

The VM can be configured in two main flavours, interpreter-only or interpreter+JIT \footnote{combining an interpreter and a JIT has performance and complexity advantages, as well as complexity disadvantages. But discussion is beyond the scope of this paper.}. Although the version with the JIT is the most widely used in production, the interpreter version is convenient for development purposes and essential on devices that outlaw JITs. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.

@@

\paragraph{Snapshots.}
Smalltalk is an object \emph{system}, rather than a language. The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap. 
%A Smalltalk system is started by loading a snapshot file into a VM. The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes. At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.  
%A new snapshot can be made during or at the end of the session.
When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods to and from the class hierarchy. Programming may be done live, as the application under development is running. %, for example creating or revising methods in the debugger, and continuing. 
For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
. A new snapshot can be made during or at the end of the development session.

Snapshots can also be used to avoid long start-up times when fixing specific bugs. The VM can be run to a point where the bug is about to manifest and a snapshot taken. Then multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.

@@

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the production VM and its simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left, low address, side, we can first find Text, the section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, including the VM's global variables.

At higher addresses we find the beginning of the memory managed by the VM. At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional section exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector. Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.

Usually at a very high address we find the C stack. Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone. Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.

\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array. References between objects are indices into the byte array instead of pointers. All the Slang variables, normally translated to C variables, are instance variables of Smalltalk objects. When required, they use specific wrapper classes, such as \emph{CArrayAccessor}, over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.). The Slang code is executed as Smalltalk code. The Smalltalk stack zone is represented in \openSmalltalkVM as a doubly linked cycle of stack pages which are maintained by the VM. In the interpreter-only simulator, the stack zone is a Smalltalk \emph{Array} object. In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, but with the JIT the Smalltalk stack must be accessible to the processor simulator.}.

@@

The Slang code is implemented in multiple Smalltalk classes, to organise the code, to add modularity through polymorphism, and reuse through inheritance. For example, the \emph{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compaction one. For production, at Slang-to-C compilation time, all the code is compiled into a single C file\footnote{one C file contains the interpreter and memory manager, with one JIT file per back end}. All the modularity is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be reused for debugging purposes. Still in our example, the \emph{AbstractCompactor} class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

@@

In addition to the interpreter simulator, simulating the JIT requires simulating execution of the machine code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. When the JIT is enabled, the start of the byte array representing the memory is used to hold the machine code it generates at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that machine code can be executed safely. All but the MIPSLE simulators are accessed through %Smalltalk primitives.  These are effectively a combination of an 
FFI calls and surrounding glue code.
Each processor simulator is invoked via calls that supply the byte array as a parameter, and effectively the byte array is the processor simulator's memory, while the processor simulator contains its register state which is accessed via Smalltalk's support for external memory access. \footnote{http://www.squeaksource.com/Alien}
% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how machine code is interfaced to the rest of the simulation. %, and the debugging facilities for machine code that are available.

@@

The interpreter has \emph{framePointer}, \emph{stackPointer} and \emph{instructionPointer} variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the \emph{stackPointer} and \emph{framePointer} variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text; all have fixed addresses. Referencing these variables and invoking these runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference each variable or routine, and uses this as the key in the relevant map, creating and storing a suitable closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will return an error from the FFI call.  Support code analyses the failure and raises a \emph{ProcessorException} which contains all necessary details, the type, pc, next pc and so on.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk \emph{ProcessorException} object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is always invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, \emph{handleSimulationTrap:}, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, \emph{ProcessorException}, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.

@@

There is a small polymorphic scheme for breakpoints.  Booleans, integers, and arrays understand \emph{isBreakpointFor: pc}.  So if the \emph{breakPC} variable is \emph{false} there is no breakpoint.  If \emph{breakPC} is \emph{true} then any pc is a potential breakpoint.  If it is an integer then that pc is a potential breakpoint.  If the pc is an \emph{Array} then the pc is a breakpoint if any value in the array answers true to \emph{isBreakpointFor:}. The scheme could trivially be extended to include intervals. If \emph{breakPC isBreakpointFor: pc} is \emph{true} then the \emph{breakBlock} is sent \emph{shouldStopIfAtPC: pc}.  booleans and closures understand \emph{shouldStopIfAtPC:}.  \emph{true} and \emph{false} make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.  For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.  It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session).

@@

Simulating the whole VM requires going through the entire Smalltalk system start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. In the simulator start-up takes around 15 seconds on a recent MacBook Pro. While developing the JIT, this start-up time may still be too long and degrade the live programming experience to an edit-compile-run cycle, which we prefer to avoid. In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part. This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to erratic behavior earlier than expected.

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows the JIT to be invoked as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code, decorates the disassembly and then dumps the output in a text window.

@@

\paragraph{Decorated Disassembly.}
Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in that disassembly and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the disassembly much more readable than that displayed in a typical machine oriented debugger.  For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64, in this case from an in-image compilation.

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, a check which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated as the machine code zone is compacted to make space for new code by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen. %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

@@

In this section we describe some memorable and representative personal experiences using the simulation infrastructure to develop the VM. The different subsections describe various tasks we undertook and in each case we emphasise how we used the simulator and how it helped us perform each given task.

\subsection{Initial JIT development and new back ends}
When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable rhythm.  As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the \emph{MessageNotUnderstood} exception. The programmer would implement the body of the method and execution would continue.  Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good (somewhat like lemming debugging below).  As a new method was encountered and JITted yet more templates would be encountered.  Similarly as the abstract machine instruction to concrete machine instruction mappings were added these could be implemented in the debugger, an experience repeated when new back ends were added. Much of the JIT was written in this interactive live style.  This style encourages tool creation because one can create, extend and polish tools in the middle of debugging sessions, enhancing them to make sense of ones current predicament.  Immediate feedback makes tool investment cheap and cheerful.

@@

Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.  Therefore one is often stuck trying to create a reproducible case in a system that is essentially chaotic: the slightest change in the heap might change the bug; any variability in timing or user input can result in a different heap and hence in the bug morphing or going into hiding.

Here, simulation comes to the rescue. Because the simulation is simply an object graph it is easy to copy the entire simulation. Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to first run the collection in the copy, performing a leak check afterwards (if enabled).  Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.  We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging%\footnote{http://www.romainbrasier.fr/404/}
.  One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original. There is no need to create a reproducible case for bugs encountered during simulation\footnote{Alas not for bugs in the production VM.  But Lemming Debugging \emph{could} be implemented in C using fork to create the copy in a child process, continuing only if the child successfully completes its GC.  Our leak checker works both in Simulation and the production VM.}; the reproducible case is at hand.  This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.

Recently, to evaluate some new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results. Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative. The whole implementation was done in the simulator, and only when it was working there was it compiled to C. Using this process, the compiled C code worked first time.

@@

@@

One of the first analyses we ran was on the machine code zone. We stopped the simulation when the machine code zone reached 1Mb. We then iterated over it and investigated what was in it. As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (they have never been executed).

@@

Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% are megamorphic.
